---
title: "Storm Data Analysis and Exploration"
author: "Panayiotis"
date: "8/19/2020"
output: 
    html_document: 
        keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary
In this document we will explore the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. The NOAA database tracks characteristics 
of major storms and weather events in the United States, including when and where 
they occur, as well as estimates of any fatalities, injuries, and property damage.
The recorded events in the NOAA database took place between 1950 and 2011. In the 
earlier years of the database there are generally fewer events recorded, most 
likely due to a lack of good records, but more recent years should be considered 
more complete. By analyzing the data, we found that tornadoes are responsible
for the majority of the deaths (37.2%) and injuries (65%), compared to all the 
other events that are included in the database. As regards the costs of the 
damages, we discovered that floods had the greatest economic (most expensive) 
consequences, surpassing all other events with a significant 31.5% of the total 
damage costs. However, to make any suggestions we would need to analyse more 
recent events (2011-2020) as many things have changed since 2011.

## Requirements:
Load necessary libraries:
```{r load_libraries}
library(data.table)
library(tidyr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(R.utils)
```

Display session info:
```{r session_info, cache=TRUE}
sessionInfo()
```

## Data Processing
1. Read the data from the url, no need to download it.
2. Convert dataframe to tibble for easier handling.
3. Convert the BGN_DATE to date object.
4. Create a new column with the date and time using BGN_DATE and BGN_TIME columns.
```{r get_data, cache=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
storm_data <- fread(url)
storm_data <- as_tibble(storm_data)
storm_data$BGN_DATE <- as.Date(storm_data$BGN_DATE, format="%m/%d/%Y")
storm_data$Time <- as_datetime(paste(storm_data$BGN_DATE, storm_data$BGN_TIME), 
                               format="%Y-%m-%d %H%M")
```

## Exploratory Data Analysis (EDA)
Let's take a look at the data:
```{r eda, cache=TRUE}
str(storm_data)
summary(storm_data)
```

## Results
### Harmful events
To find which events have the most harmful impact on the population we have to
use the FATALITIES and INJURIES columns.
```{r harmful_events, cache=TRUE}
harmful_events <- storm_data %>% 
    select(EVTYPE, FATALITIES, INJURIES) %>% 
    group_by(EVTYPE) %>% 
    mutate(Total_Fatal=sum(FATALITIES), Total_Inj=sum(INJURIES)) %>% 
    select(EVTYPE, Total_Fatal, Total_Inj) %>% 
    distinct() 
```

Now let's calculate the sum and the percentage of fatalities and injuries, in 
order to better understand the impact of each event. This is also a form of 
normalization that is applied on the total values.
```{r calc_percentages, cache=TRUE}
sum_fatal <- sum(harmful_events$Total_Fatal)
fatal_events <- harmful_events %>% 
    arrange(desc(Total_Fatal), desc(Total_Inj)) %>%
    mutate(Prop_Fatal=Total_Fatal / sum_fatal * 100)

sum_inj <- sum(harmful_events$Total_Inj)
injury_events <- harmful_events %>% 
    arrange(desc(Total_Inj), desc(Total_Fatal)) %>%
    mutate(Prop_Inj=Total_Inj / sum_inj * 100)
```

It's time to plot our data! We will use only the top 10 top events according to
fatalities and injuries, since these are the most harmful events and plotting 
everything would be too messy and pointless.
```{r plot_1}
fatal_events$EVTYPE <- with(fatal_events, reorder(EVTYPE, -Prop_Fatal))
ggplot(fatal_events[1:10, ], aes(EVTYPE, Prop_Fatal, fill=EVTYPE)) + 
    geom_bar(stat="identity") + theme_minimal() + 
    theme(legend.position="none", axis.title.y=element_blank()) +
    labs(y="Total Fatalities (%)", title="Top 10 Most Fatal Events in USA") +
    coord_flip() 
```

The above plot shows the top 10 most fatal event in USA based on the total
fatalities. In particular, this figure illustrates the percentage of the total
fatalities that corresponds to each event. As we can see, tornado is clearly the
one that caused the most deaths with a whouping 37.19%, excessive heat follows 
with 12.57% and then we have flash flood and heat at 6.46% and 6.19%, respectively.

Below you can see a table with all the corresponding values.
```{r display_fatalities}
fatal_events[1:10, c("EVTYPE", "Total_Fatal", "Prop_Fatal")]
```

```{r plot_2}
injury_events$EVTYPE <- with(injury_events, reorder(EVTYPE, -Prop_Inj))
ggplot(injury_events[1:10, ], aes(EVTYPE, Prop_Inj, fill=EVTYPE)) + 
    geom_bar(stat="identity") + theme_minimal() + 
    theme(legend.position="none", axis.title.y=element_blank()) +
    labs(y="Total Injuries (%)", title="Top 10 Events with the Most Injuries in USA") +
    coord_flip()
```

The above barplot illustrates the top 10 events that caused the most injuries in
USE. The values were normalized in order to show the proportions of the total
injuries that took place for each event. It is obvious that tornado was responsible 
for the majority of injuries, which corresponds to approximately 65% of the total
injuries. TSTM wind, flood and excessive heat follow, with 4.95%, 4.83% and 4.64%,
respectively.

The corresponding values can be found below.
```{r display_injuries}
injury_events[1:10, c("EVTYPE", "Total_Inj", "Prop_Inj")]
```

#### Conclusions about harmful events:
According the above barplots, it seems that the most harmful event in United States
is the Tornado, as it is responsible for about 37.2% of fatalities and the massive 
65% of the total injuries.

### Economic consequences
Now let's see which events have the most economic consequences. To do that we need
to check the PROPDMG and PROPDMGEXP columns to get the propert damage expenses and
the CROPDMG and CROPDMGEXP to get the crop damage expenses. Before adding these
two types of expenses, we must fist calculate the actual values. 

#### Important assumptions:
We assume that "+", "-", "-", "?", "0", and "" have exponent 0 which means that 
the actual value will be equal to the PROPDMG or CROPDMG. For "H", "K", "M", "B" 
we assume that they correspon to hundreds, thousands, millions and billions, 
respectively. For the numbers between 1-8 we assume that these represent the 
exponent with base 10.
```{r calc_expenses, cache=TRUE}
calc_value <- function(multiplier, exponent){ 
    v <- sapply(seq_along(exponent), function(x) switch (exponent[x],
        "H" = 10^2,
        "K" = 10^3,
        "M" = 10^6,
        "B" = 10^9,
        "1" = 10,
        "2" = 10^2,
        "3" = 10^3,
        "4" = 10^4,
        "5" = 10^5,
        "6" = 10^6,
        "7" = 10^7,
        "8" = 10^8,
        1 
    ))
    return(multiplier * v)
}
expenses <- storm_data %>%
    select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>% 
    group_by(EVTYPE) %>%
    mutate(prop_expenses=calc_value(PROPDMG, PROPDMGEXP), 
           crop_expenses=calc_value(CROPDMG, CROPDMGEXP),
           Total_Expenses=prop_expenses + crop_expenses,
           Cum_Expenses=sum(Total_Expenses)) %>%
    select(EVTYPE, Cum_Expenses) %>% 
    distinct() %>%
    arrange(desc(Cum_Expenses))
```

It is better to normalise the expenses values so we will calculate the percentages
(proportions) as before.
```{r calc_proportions, cache=TRUE}
sum_expenses <- sum(expenses$Cum_Expenses)
costly_events <- expenses %>% 
    mutate(Prop_Expenses=Cum_Expenses / sum_expenses * 100)
```

Let's create a plot with the top 10 events with the greatest economic consequences.
```{r plot_3}
costly_events$EVTYPE <- with(costly_events, reorder(EVTYPE, -Prop_Expenses))
ggplot(costly_events[1:10, ], aes(EVTYPE, Prop_Expenses, fill=EVTYPE)) + 
    geom_bar(stat="identity") + theme_minimal() + 
    theme(axis.title.x=element_blank(), 
          axis.ticks.x=element_blank(), axis.text.x=element_blank()) +
    labs(y="Total Expenses (%)", title="Top 10 Events with the Greatest Economic Consequences in USA")
```

The above barplot shows the proportions of the total expenses for each type of event.
More specifically, it shows the top 10 events that had the greatest economic consequences,
in terms of property and crops damage costs. Clearly, flood caused the most damages
as it was responsible for approximately 31.50% of the total damages in USA. 
Hurricane/Typhoon and tornado follow with 15.07% and 12.02%, respectively.

Below you can see the corresponding values used in the barplot.
```{r display_costs}
costly_events[1:10, ] 
```

#### Conclusions about economic consequences:
Based on the previous plot, it seems that the flood caused the most damages in USA,
with a 31.5%, which is almost one third of the total damages caused by all
types of events. With this finding, we could suggest that further measures should
be taken to protect the properties and crops from flood events, however, further
analysis should be made to check whether this finding applies for the current decade
or only for previous decades.

### Cleanup
Now that we are done let's clean our session:
```{r cleanup}
rm(url)
rm(storm_data)
rm(harmful_events)
rm(fatal_events)
rm(injury_events)
rm(expenses)
rm(costly_events)
rm(sum_expenses)
rm(sum_fatal)
rm(sum_inj)
rm(calc_value)
detach("package:data.table", unload=TRUE)
detach("package:tidyr", unload=TRUE)
detach("package:ggplot2", unload=TRUE)
detach("package:dplyr", unload=TRUE)
detach("package:lubridate", unload=TRUE)
detach("package:R.utils", unload=TRUE)
```

Let's check if we cleaned everything:
```{r}
ls()
```
